Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-10-06 Tex/Paper.tex
Edits w/Erik

Beam driven plasma wakefield accelerators \cite{chen:1985} have the potential to offer compact linear accelerators with
high energy gradients, and have been of interest for several decades. With a relativistic charged particle drive beam
travelling through a plasma, a strong wakefield is excited, that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. The concept has
been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008, litos:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread
and emittance growth small. In the well described linear regime, valid when the beam density $n_{b}$ is much smaller
than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam.
The beam also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the
beam has been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons, gathering in a sheet around an evacuated area filled with only
ions. The ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly
with radius. This focusing force preserves emittance \cite{rosenzweig:1991}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can
be ensured, when accelerated by a proton drive beam producing quasi-linear wakefields \cite{rosenzweig:2010}. By
quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low relative
energy spread, and completely blow out the electrons left in the bubble after the beam to reach conditions that
preserves emittance. The results have importance for the preparation of AWAKE Run 2 and possibly other applications in
the quasi-linear regime.

@@

The energy carried by electron drive bunches used in previous plasma wakefield experiments have been no more than
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}.
% VKBO: Not sure "no more than" is true.
For high-energy physics application a higher total beam energy is often desired, for instance, the energy of a
high-charge electron bunch accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be
produced by the international linear collider, is $1.6\unit{kJ}$. Using electron beams as drivers a large number of
plasma stages is required to reach an energy of a $\unit{kJ}$ for the accelerated bunch. However, staging plasma
accelerators without reducing the effective gradient and spoiling the beam quality is challenging
\cite{steinke:2016,lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS
beam \cite{gschwendtner:2016}, allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of
magnitude longer than the plasma wavelengths needed for such applications, and the beam does not drive a strong wake
\cite{gschwendtner:2016}. By letting the proton beam undergo self-modulation before injecting the witness beam into the % added "the"
accelerating structure, stronger wakefields are excited. The self-modulation of the beam is produced by the transverse
fields generated by the beam acting upon itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The
modulation frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis
with a surrounding halo of defocused particles. This train of bunches resonantly drives wakefields to large amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield accelerator experiment
\cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton
beam delivered by the SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$
-- matched to the transverse size of the SPS proton beam such that $k_{pe}\sigma_{x,y,pb} = 1$, where
$k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number, $\lambda_{pe}=2\pi c/\omega_{pe}$ and
$\omega_{pe}=\left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron angular frequency.

@@

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of about $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the
electron beam present. The electron witness beam is then injected into the modulated proton beam before section two,
where it undergoes acceleration. The self-modulated proton beam does not produce a fully non-linear wakefield, and
therefore not all plasma electrons are evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius and the accelerated beam emittance is not preserved.

@@

The focus in this study is the beam loading of the wakefields driven by the proton beam. Studies of self-modulated
proton bunches show that the proton bunch evolve as it propagates through a uniform plasma \cite{lotov:2011}, but that
small variations in the plasma profile the modulation, and thus the wake fields, may be stabilised over long distances
\cite{lotov:2011, lotov:2015, caldwell:2011}. To study the witness beam evolution in a stable wake, independent of the
dynamics of the self-modulation, here we use a single, non-evolving proton bunch as driver. The proton bunch parameters
are chosen so that key features of the wake -- the plasma electron density in the wake and the longitudinal electric
field -- are the same as in the wake of a self-modulated proton beam with AWAKE baseline parameters
\cite{gschwendtner:2016}. Both the proton beam and the witness have Gaussian longitudinal charge distribution and
bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full PIC code Osiris \cite{fonseca:2002}
and with 2D cylindrical-symmetric simulations. The studies \cite{berglyd_olsen:2015, berglyd_olsen:2016} primarily
looked at beam loading, energy gain and energy spread, as well as different approaches to create a stable drive beam
structure based on previous self-modulation studies. In order to study the witness beam emittance evolution we adopt the
recently released open source version of QuickPIC \cite{huang:2006, an:2013}. QuickPIC is a fully relativistic 3D
quasi-static PIC code. It does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited tool to study emittance preservation.  All simulation results in
this paper are obtained using QuickPIC open source.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} Plasma density in grey with the drive beam (blue) and the witness beam (red)
        superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis. Beams move to the left.}
\end{figure}

@@

The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void
of plasma electrons. When the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second
plasma stage (figure \ref{Fig:AWAKER2}), the peak electric field is expected to be about $500\funit{MV}{m}$. The plasma
electron density are only depleted to around $65\%$ of nominal plasma density at the point where we inject the electron
beam \cite{awake_collaboration:2016}.  The plasma electron density depletion and the peak field is replicated using a
single bunch with $1.46\nexp{10}$ protons (or $2.34\unit{nC}$), a length $\sigma_{z} = 40\unit{\mu m}$ ($7\unit{kA}$),
and a transverse size $\sigma_{x,y,pb} = 200\unit{\mu m}$. The peak density is $0.83\cdot n_{0}$ results in a
quasi-linear wake. To avoid transverse evolution of the proton driver, emulating the stable propagation of the
self-modulated proton beam \cite{lotov:2011, lotov:2015, caldwell:2011} we freeze the transverse evolution of the
equivalent proton bunch by increasing the particles mass by six orders of magnitude.

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}
as well as to its de-phasing with respect to the wakefields when at low energy. To eliminate de-phasing of the witness
beam, the initial beam energy is set such that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$.
A lower initial energy is likely to be sufficient for AWAKE Run 2 injection.
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement,
%     no details in \cite{berglyd_olsen:2015}.

Eq. (\ref{EQ:Matched}) yields a transverse size of $\sigma_{x,y,eb}$ of $5.25\unit{\mu m}$, narrow compared to the drive
beam $\sigma_{x,y,pb} = 200\unit{\mu m}$. The bunch length was set to $\sigma_{z}=60\unit{\mu m}$, based on earlier beam
loading studies \cite{berglyd_olsen:2016}. The charge is adjusted to $100\unit{pC}$ for optimal beam loading, as
discussed in the next section. We refer to the defined drive beam and witness beam parameter set as the
\emph{base case}.
% EA: the key figure, figure 2, was not discussed in the text until very far down!
%     I therefore added (text could possibly be improved)
Figure \ref{Fig:PlasmaDenTWake} shows the two beams from QuickPIC simulations of the initial time step, for the base
case parameters; the proton bunch in blue, the trailing electron beam in red and the plasma electron density in grey.

@@

The relatively small size of the witness beam puts constraints on the transverse cell size and number in the
simulations: small size to resolve the small electron beam and large number of to resolve the large proton beam and the
wakefields. % This sounds really bad? Should be improved.
We use a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal grid cells for
the simulations presented in section \ref{S:BL}. The witness beam was simulated with $16.8\nexp{6}$ non-weighted
particles.
%  ***NN*** plasma particles were simulated per grid cell.
% EA: Veronica to fill in NN above
% VO: QuickPIC doesn't specify particles per cell. Only in Osiris, This number has no fixed value, just a range from
%     zero to whatever is the density at the peak of the Gaussian distribution.

@@

The witness beam generates its own wakefield which loads the $E_{z}$ field generated by the drive beam. With an ideally
shaped electron beam charge profile it is possible to optimally load the field in such a way that the accelerating field
is constant along the beam \cite{katsouleas:1987, tzoufras:2009}. Gaussian beams, as assumed in these studies, cannot
completely flatten the electric field in the tails of the charge distribution, and our base case beam therefore has a
tail in energy both at the front and the back of the beam, as illustrated in figure \ref{Fig:BeamPS}. The bulk of the
beam, however, sees a relatively flat field.
% EA: I rewrote this to the sentence after the comments
%we can acceleratoreHowever, with such a narrow beam when matched beam to the plasma density (equation
%\ref{EQ:MatchedB}), we are limited on how much total charge we can accelerate without reaching charge densities that
%will overload the field.
%  EA: shall we say the below?  The sentence asks for more detailed studies of the dependence of the loading of the transverse fields.
%The amounts of charge required to load the field depends both on the witness bunch length as well as the transverse size.

The electron beam density is $n_{eb} \approx 35\times n_{0}$. This means that the witness beam's own wakefield is in the
fully non-linear regime, where the space charge force is sufficient to blow-out all plasma electrons resulting in the
formation of an ion column (see figure \ref{Fig:BeamLoading}, bottom). This ion column, as is well known
\cite{rosenzweig:1991}, provides a linear focusing force on the part the electron beam within the column, and therefore
prevents emittance growth for this part of the beam.  This effect is shown for our base case in figure
\ref{Fig:PlasmaDenTWake}. The focusing field has a gradient of $20\unit{kT/m}$ near the beam axis, corresponding to the
matched field gradient.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
% EA: I cut the line below, we are still discussing the base case
%For the majority of the cases we studied, that maintained a stable accelerating structure, 
 %PMDue to the ion column focusing, 
Simulation results show that about $70-80\%$ of the electron beam longitudinal slices retain their initial emittance
after the propagation in the plasma. Figure \ref{Fig:BeamEmitt} shows the slice emittance along the beam for the base
case, sampled after propagating through $0$, $4$, $40$ and $100\unit{m}$ of plasma. Emittance growth mainly occurs in
first few meters, and no significant emittance growth is observed after this for propagation lengths up to
$100\unit{m}$. The head of the beam does not benefit from the full ion column focusing, but since the proton beam
creates a quasi-linear wake, the emittance of the head of the beam still stabilizes after some time. For the
$100\unit{m}$ simulation, the drive beam energy was increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing, as
de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$ after about $50\unit{m}$.

%~ Our configuration also shows some robustness to small electron beam transverse offsets as the bubble generated by the
%~ witness beam wakefield follows the beam itself.
So far we have considered a witness beam injected on the axis of the proton bunch. We now briefly examine the case of
injection of a witness beam with an offset with respect to the proton bunch axis. Since the witness beam creates its own
plasma bubble, the emittance of the part of the beam inside the bubble is not affected by small transverse offsets of
the witness beam with respect to the proton beam axis. This is illustrated in figure \ref{Fig:BeamEmitt}, right, for an
electron beam offset of $1\sigma_{x,eb}$. Emittance preservation for small offsets is an added benefit of this
accelerating regime, and may ease transverse injection tolerances. The head of the beam experiences a 
%larger initial emittance growth (see figure \ref{Fig:BeamFilament}a-f),
%PM
%THERE IS NO EMITTANCE SHOWN OR NOTHING TO COMPARE TO ... WHAT TO DO????
% EA: one way out, can compare emittance growth using fig 5 instead, as follows (alternatively, a new set of figures must be created)
% VKBO: How does the phase space plots *NOT* show emittance? 
larger initial emittance growth than for the on-axis case (compare figure \ref{Fig:BeamEmitt}, left, to
\ref{Fig:BeamEmitt}, right), however, also for the head the emittance growth ends after the first few meters.
Figure \ref{Fig:BeamFilament}a-c show the phase space of the head of the electron beam after $0$, $1.0$ and
$2.5\unit{m}$, while figure \ref{Fig:BeamFilament}d-f show the phase space of the trailing part of the beam. The
centroid oscillations of the head and the trailing part are shown in figure \ref{Fig:BeamFilament}g. This effect of a
transverse offset is likely to be greater for larger offsets as the beam oscillates around the axis of the drive beam
wakefield.
% EA: would be good to include a statement of how large the offsets can be before when the electron beam starts to break
%     up / not preserve emittance.

The transverse beam size within the bubble, where normalized emittance is preserved, follows the evolution given by
equation \ref{EQ:Matched},i .e., evolves to stay matched. The on-axis density of the electron beam, as a result,
increases as its gamma factor increases and its transverse size decreases. The effect can be seen in figure
\ref{Fig:BeamFilament}h. This has the potential to cause overloading of the field. However, for the base case no
significant overloading is observed.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQuality}
    \caption{\label{Fig:BeamQ} Ratio of total beam charge $\widetilde{Q}$ wit h an emittance growth
        $\Delta\epsilon \leq 5\%$ as a function of initial beam charge (blue symbols, lines), and relative energy spread
        of the accepted charge (red symbols, lines), after $4\unit{m}$ of plasma and with an initial emittance
        $\epsilon_{\mathrm{N},0}=2\unit{\mu m}$. These are shown for four different $\sigma_{z}$ from $40\unit{\mu m}$
        to $100\unit{\mu m}$. The detailed studies presented in beam loading section correspond to the square marked
        lines at $100\unit{pC}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQualityAbs}
    \caption{\label{Fig:BeamQAbs} Total beam charge $\widetilde{Q}$ with an emittance growth $\Delta\epsilon \leq 5\%${}
        as a function of initial beam charge (blue symbols, lines), and final energy (red symbols, lines), after
        $4\unit{m}$ of plasma and with an initial emittance $\epsilon_{\mathrm{N},0}=2\unit{\mu m}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQualityEmittance}
    \caption{\label{Fig:BeamQEmit} Ratio of total beam charge $\widetilde{Q}$ with an emittance growth
        $\Delta\epsilon \leq 5\%$ as a function of initial emittance (blue symbols, lines), and relative energy spread
        of the accepted charge (red symbols, lines), after $4\unit{m}$ of plasma.}
\end{figure}

@@

The beam loading and blow out properties of the electron beam depend on a large number of parameters, including
longitudinal profile, transverse profile as well as relative phasing of the proton and the electron beams. We present a
limited parameter study aimed at guiding beam parameter choices for AWAKE Run 2. For electron beam to be externally
injected in AWAKE Run 2 (see figure \ref{Fig:AWAKER2}) it is desired to maximize the energy gain, minimize the energy
spread, maximise the charge to be accelerated, and minimise the emittance growth \cite{adli:2016}. In addition, the
bunch length should be such that it is possible to generate and transport the bunch using a compact electron injector
\cite{adli:2016}. We investigate the interdependence of these parameters in simulation by varying the electron bunch
length, its charge, and initial emittance. The results were quantified in terms of how much of the initial charge
retained its initial emittance within a $5\%$ margin. For these parameter scans we used a transverse grid cell size of
$2.34\unit{\mu m}$, and let the beams propagate through $4\unit{m}$ of plasma.

>>>>

Beam driven plasma wakefield accelerators \cite{chen:1985} have the potential to offer compact linear accelerators with
high energy gradients, and have been of interest for several decades. With a relativistic charged particle drive beam
travelling through a plasma, a strong wakefield is excited that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. The concept has
been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008, litos:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread
and emittance growth small. In the well described linear regime, valid when the beam density $n_{b}$ is much smaller
than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam.
The beam also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the
beam has been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons, gathering in a sheet around an evacuated area filled with only
ions. The ions, assumed stationary, form a uniform density ion channel creating a focusing force that varies linearly
with radius. This focusing force preserves emittance \cite{rosenzweig:1991}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can
be ensured when accelerated by a proton drive beam producing quasi-linear wakefields \cite{rosenzweig:2010}. By
quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low relative
energy spread, and completely blow out the electrons left in the bubble after the beam, to reach conditions that
preserves emittance. The results have importance for the preparation of AWAKE Run 2 and possibly other applications in
the quasi-linear regime.

@@

The energy carried by electron drive beams used in previous plasma wakefield experiments have been on the order of
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$ \cite{blumenfeld:2007, caldwell:2009}.
For high-energy physics application a higher total beam energy is often desired, for instance, the energy of a
high-charge electron beam accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be
produced by the international linear collider, is $1.6\unit{kJ}$. Using electron beams as drivers a large number of
plasma stages is required to reach an energy of a $\unit{kJ}$ for the accelerated beam. However, staging plasma
accelerators without reducing the effective gradient and spoiling the beam quality is challenging
\cite{steinke:2016, lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS
beam \cite{gschwendtner:2016}, allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of
magnitude longer than the plasma wavelengths needed for such applications, and it does not drive a strong wake
\cite{gschwendtner:2016}. By letting the proton beam undergo self-modulation before injecting the witness beam into the
accelerating structure, stronger wakefields are excited. The self-modulation is produced by the transverse fields
generated by the beam acting upon itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The
modulation frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis
with a surrounding halo of defocused particles. This train of bunches resonantly drives wakefields to large amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield accelerator experiment
\cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton
beam delivered by the SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$
-- matched to the transverse size of the SPS proton beam such that $k_{pe}\sigma_{x,y,pb} = 1$, where
$k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number, $\lambda_{pe} = 2\pi c/\omega_{pe}$, and
$\omega_{pe} = \left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron angular frequency.

@@

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of about $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the
electron beam present. The electron witness beam is then injected into the modulated proton beam before section two
where it undergoes acceleration. The self-modulated proton beam does not produce a fully non-linear wakefield, and
therefore not all plasma electrons are evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius and the accelerated beam emittance is not preserved.

@@

The focus in this study is the beam loading of the wakefields driven by the proton beam. Studies of self-modulated
proton beams show that the proton beam evolves as it propagates through a uniform plasma \cite{lotov:2011}, but that
small variations in the plasma profile the modulation, and thus the wake fields, may be stabilised over long distances
\cite{lotov:2011, lotov:2015, caldwell:2011}. To study the witness beam evolution in a stable wake, independent of the
dynamics of the self-modulation, we use a single, non-evolving proton bunch as driver. The proton beam parameters
are chosen so that key features of the wake -- the plasma electron density in the wake and the longitudinal electric
field -- are the same as in the wake of a self-modulated proton beam with AWAKE baseline parameters
\cite{gschwendtner:2016}. Both the proton beam and the witness beam have Gaussian longitudinal charge distribution and
bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full PIC code Osiris \cite{fonseca:2002}
and with 2D cylindrical-symmetric simulations. The studies \cite{berglyd_olsen:2015, berglyd_olsen:2016} primarily
looked at beam loading, energy gain and energy spread, as well as different approaches to creating a stable drive beam
structure based on previous self-modulation studies. In order to study the witness beam emittance evolution we use the
recently released open source version of QuickPIC \cite{huang:2006, an:2013}. QuickPIC is a fully relativistic 3D
quasi-static PIC code. It does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited tool to study emittance preservation. All simulation results in
this paper were obtained using QuickPIC open source.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} QuickPIC simulations of the initial time step, for the single proton drive beam
        and witness beam setup. Plasma electron density is shown in grey with the drive beam (blue) and the witness beam
        (red) superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis. Beams move to the left.}
\end{figure}

@@

The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void
of plasma electrons. When the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second
plasma section (figure \ref{Fig:AWAKER2}), the peak electric field is expected to be about $500\funit{MV}{m}$. The
plasma electron density is only depleted to around $65\%$ of nominal plasma density at the point where we inject the
electron beam \cite{awake_collaboration:2016}. The plasma electron density depletion and the peak field is replicated
using a single bunch with $1.46\nexp{10}$ protons ($2.34\unit{nC}$), a length $\sigma_{z} = 40\unit{\mu m}$
($7\unit{kA}$), and a transverse size $\sigma_{x,y,pb} = 200\unit{\mu m}$. The peak density at $0.83\cdot n_{0}$ results
in a quasi-linear wake. To avoid transverse evolution of the proton driver, emulating the stable propagation of the
self-modulated proton beam \cite{lotov:2011, lotov:2015, caldwell:2011}, we freeze the transverse evolution of the
equivalent proton bunch by increasing the particles mass by six orders of magnitude.

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}
as well as to its de-phasing with respect to the wakefields at low energy. To eliminate de-phasing of the witness beam,
the initial beam energy is set such that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$. A
lower initial energy is likely to be sufficient for AWAKE Run 2 injection.
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement,
%     no details in \cite{berglyd_olsen:2015}.

Eq. (\ref{EQ:Matched}) yields a transverse size $\sigma_{x,y,eb}$ of $5.25\unit{\mu m}$, which is narrow compared to the
drive beam $\sigma_{x,y,pb} = 200\unit{\mu m}$. The bunch length was set to $\sigma_{z} = 60\unit{\mu m}$, based on
earlier beam loading studies \cite{berglyd_olsen:2016}. The charge is adjusted to $100\unit{pC}$ for optimal beam
loading, as discussed in the next section. We refer to the defined drive beam and witness beam parameter set as the
\emph{base case}. Figure \ref{Fig:PlasmaDenTWake} shows the two beams from QuickPIC simulations of the initial time
step, for the base case parameters; the proton beam in blue, the trailing electron beam in red and the plasma electron
density in grey.

@@

The relatively small size of the witness beam puts constraints on the transverse grid cell size and number in the
simulations. We need a small size to resolve the narrow electron beam, and a large number of grid cells to resolve the
much wider proton beam and its wakefields. We use a transverse grid cell size of $1.17\unit{\mu m}$, and of
$2.34\unit{\mu m}$ for the longitudinal grid cells for the simulations presented in section \ref{S:BL}. The witness beam
was simulated with $16.8\nexp{6}$ non-weighted particles, the drive beam with $2,1\nexp{6}$ non-weighted particles, and
the plasma electrons with $1024 \times 1024$ weighted particles per transverse slice.

@@

The witness beam generates its own wakefield which loads the $E_{z}$-field generated by the drive beam. With an ideally
shaped electron beam charge profile it is possible to optimally load the field in such a way that the accelerating field
is constant along the beam \cite{katsouleas:1987, tzoufras:2009}. Gaussian beams, as assumed in these studies, cannot
completely flatten the electric field in the tails of the charge distribution, and our base case beam therefore has a
tail in energy both at the front and the back of the beam, as illustrated in figure \ref{Fig:BeamPS}. The bulk of the
beam, however, sees a relatively flat field.
% EA: I rewrote this to the sentence after the comments
%we can acceleratoreHowever, with such a narrow beam when matched beam to the plasma density (equation
%\ref{EQ:MatchedB}), we are limited on how much total charge we can accelerate without reaching charge densities that
%will overload the field.
%  EA: shall we say the below?  The sentence asks for more detailed studies of the dependence of the loading of the transverse fields.
%The amounts of charge required to load the field depends both on the witness bunch length as well as the transverse size.

The initial electron beam density is $n_{eb} \approx 35\times n_{0}$. This means that the witness beam's own wakefield
is in the fully non-linear regime, where the space charge force is sufficient to blow-out all plasma electrons resulting
in the formation of an ion column (see figure \ref{Fig:BeamLoading}, bottom). This ion column, as is well known
\cite{rosenzweig:1991}, provides a linear focusing force on the part the electron beam within the column, and therefore
prevents emittance growth for this part of the beam. This effect is shown for our base case in figure
\ref{Fig:PlasmaDenTWake}. The focusing field has a gradient of $20\unit{kT/m}$ near the beam axis, corresponding to the
matched field gradient.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
% EA: I cut the line below, we are still discussing the base case
%For the majority of the cases we studied, that maintained a stable accelerating structure, 
 %PMDue to the ion column focusing,

Figure \ref{Fig:BeamEmitt} shows the slice emittance along the beam for the base case, sampled after propagating through
$0$, $4$, $40$ and $100\unit{m}$ of plasma. We define emittance of a slice as preserved if the growth is less than
$5\%$, and $\widetilde{Q}$ as the sum charge of the slices for which the emittance is preserved. Simulation results show
that $\widetilde{Q}/Q = 73\%$ of the electron beam longitudinal slices retain their initial emittance after the
propagation in the plasma. Emittance growth mainly occurs in the first few meters, and no significant emittance growth
is observed after this for propagation lengths up to $100\unit{m}$. The head of the beam does not benefit from the full
ion column focusing, but since the proton beam creates a quasi-linear wake, the emittance of the head of the beam still
stabilizes after some time. For the $100\unit{m}$ simulation, the drive beam energy was increased to $7\unit{TeV}$ (LHC
energy) to prevent de-phasing, as de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$
after about $50\unit{m}$.

%~ Our configuration also shows some robustness to small electron beam transverse offsets as the bubble generated by the
%~ witness beam wakefield follows the beam itself.
So far we have considered a witness beam injected on the axis of the proton beam. We now briefly examine the case of
injection of a witness beam with an offset with respect to the proton beam axis. Since the witness beam creates its own
plasma bubble, the emittance of the part of the beam inside the bubble is not affected by small transverse offsets of
the witness beam with respect to the proton beam axis. This is illustrated in figure \ref{Fig:BeamEmitt}, right, for an
electron beam offset of $1\sigma_{x,eb}$. Emittance preservation for small offsets is an added benefit of this
accelerating regime, and may ease transverse injection tolerances. The head of the beam experiences a 
%larger initial emittance growth (see figure \ref{Fig:BeamFilament}a-f),
%PM
%THERE IS NO EMITTANCE SHOWN OR NOTHING TO COMPARE TO ... WHAT TO DO????
% EA: one way out, can compare emittance growth using fig 5 instead, as follows (alternatively, a new set of figures must be created)
% VKBO: How does the phase space plots *NOT* show emittance? 
larger initial emittance growth than for the on-axis case (compare figure \ref{Fig:BeamEmitt}, left, to
\ref{Fig:BeamEmitt}, right). However, also for the head of the beam the emittance growth ends after the first few
metres. Figure \ref{Fig:BeamFilament}a-c show the phase space of the head of the electron beam after $0$, $1.0$ and
$2.5\unit{m}$, while figure \ref{Fig:BeamFilament}d-f show the phase space of the trailing part of the beam. The
centroid oscillations of the head and the trailing part are shown in figure \ref{Fig:BeamFilament}g. This effect of a
transverse offset is likely to be greater for larger offsets as the beam oscillates around the axis of the drive beam
wakefield.
% EA: would be good to include a statement of how large the offsets can be before when the electron beam starts to break
%     up / not preserve emittance.

The transverse beam size within the bubble, where normalized emittance is preserved, follows the evolution given by
equation \ref{EQ:Matched}, i.e., evolves to stay matched. The on-axis density of the electron beam, as a result,
increases as its gamma factor increases and its transverse size decreases. This effect can be seen in figure
\ref{Fig:BeamFilament}h. This has the potential to cause overloading of the field. However, for the base case no
significant overloading is observed.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQuality}
    \caption{\label{Fig:BeamQ} Ratio of beam charge with emittance preserved, $\widetilde{Q}/Q$, as a function of initial
        beam charge (blue symbols, lines), and relative energy spread of the accepted charge (red symbols, lines), after
        $4\unit{m}$ of plasma and with an initial emittance
        $\epsilon_{\mathrm{N},0}=2\unit{\mu m}$. These are shown for four different $\sigma_{z}$ from $40\unit{\mu m}$
        to $100\unit{\mu m}$. The detailed studies presented in beam loading section correspond to the square marked
        lines at $100\unit{pC}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQualityAbs}
    \caption{\label{Fig:BeamQAbs} Beam charge with emittance preserved, $\widetilde{Q}$,
        as a function of initial beam charge (blue symbols, lines), and final energy (red symbols, lines), after
        $4\unit{m}$ of plasma and with an initial emittance $\epsilon_{\mathrm{N},0}=2\unit{\mu m}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamQualityEmittance}
    \caption{\label{Fig:BeamQEmit} Ratio of beam charge with emittance preserved, $\widetilde{Q}/Q$,
        as a function of initial emittance (blue symbols, lines), and relative energy spread
        of the accepted charge (red symbols, lines), after $4\unit{m}$ of plasma.}
\end{figure}

@@

The beam loading and blow out properties of the electron beam depend on a large number of parameters, including
longitudinal profile, transverse profile as well as relative phasing of the proton and the electron beams. We present a
limited parameter study aimed at guiding beam parameter choices for AWAKE Run 2. For an electron beam to be externally
injected in AWAKE Run 2 (see figure \ref{Fig:AWAKER2}) it is desired to maximise the energy gain, minimise the energy
spread, maximise the charge to be accelerated, and minimise the emittance growth \cite{adli:2016}. In addition, the
beam length should be such that it is possible to generate and transport the beam using a compact electron injector
\cite{adli:2016}. We investigate the interdependence of these parameters in simulation by varying the electron beam
length, its charge, and initial emittance. The results are quantified in terms of how much of the beam retains its
initial emittance. For these parameter scans we used a transverse grid cell size of $2.34\unit{\mu m}$, and let the
beams propagate through $4\unit{m}$ of plasma.
